---
title: "notebook_exam_statistical_data_analysis"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: false

# load important libraries 
library(here)
library(tidyverse)

```

# Introduction

The aim of this notebook is to explain the scripts found in the folder `02_new_script2/`.\
Scripts can be divided into three categories:

- **base functions**: small scripts that generate simple functions (a wise man once said: "if you have to do it more than once, instead write a script!").\
  The base functions are:

  - `datasummary.R`

  - `filt_data.R`

  - `pseudocount.R`

  - `generate_matrix_factors.R`

  - `pielou_ind.R`

- **orchestrators**: these are slightly more complex scripts than the base functions; their job is to use base functions to generate something more complex.\
  The orchestrators are:

  - `clr_pearson.R`

  - `NorTa_simulation.R`

  - `data_sim_ph_driven.R`

- **standalone analysis**: using the previous functions, these scripts generate the analyses that are important for this project. The standalone analyses are:

  - `compositional_bias_D_P.R`

  - `ph_pielou_confrontation.R`

Let's start by looking at the **base functions** scripts:

## Base functions

<clever text here>

### datasummary.R

*What it does:* defines the function `datasum()`, which prints a compact summary of any numeric dataset (tibble, data.frame, list, or matrix).\
The summary includes:

- dimensions (rows, columns);
- basic statistics: minimum, maximum, mean, median, standard deviation, skewness;
- zero structure: total number of zeroes and their percentage;
- for square numeric matrices only: determinant and minimum eigenvalue (useful to check properties of correlation/covariance matrices).

*Why is it here:* during the project I felt the need for a quick, consistent way to inspect matrices and tables (a wise man once said: "if you have to do it more than once, instead write a script about it").\
`datasum()` centralises these checks in a single reusable function, which is particularly handy for compositional and metagenomic data, where zeroes and matrix properties matter for statistical analyses.

*How it works:* the function `datasum()`:  

- checks if `x` is a square numeric matrix and, if so and if there are no `NA`/infinite values, computes the determinant and minimum eigenvalue; 

- coerces the input to a tibble (converting lists to data frames if needed);  

- computes min, max, mean, median, standard deviation, and skewness (third standardized moment);  

- counts zeroes and their percentage;  

- prints all these quantities with a short label derived from the object name;  

- returns them invisibly as a named list, so they can be reused programmatically.

```{r}
# brings the datasummary script into scope
source(
  list.files(
    path = here(),
    pattern = "^datasummary\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# generate a matrix that will be used as example
example_matrix <- matrix(c(22, 3, 5, 7, 9, 10, 0, 0, 1),
                         nrow = 3,
                         ncol = 3,
                         dimnames = list(NULL, c("A", "B", "C")))

# print the matrix for confrontation
print(example_matrix)

# apply the datasum function from the scoped script
datasum(example_matrix)

```

### filt_data.R

*What it does:* filters a metagenomic OTU count table (samples × OTUs) by prevalence and abundance, and shows a before/after summary using `datasum()`.

*Why is it here:* metagenomic data (like HMP2 used for this project) are compositional and very sparse.\
Before any downstream analysis, it is important to remove rare and low-abundance OTUs that mostly add noise and can distort statistical results.

*How it works:* the script defines the function `filt_data()`, which:

- prints a summary of the data **before** filtering using `datasum()`;
- asks the user for a **prevalence threshold** (a number between 0 and 1), i.e. the minimum fraction of samples in which an OTU must be present (non-zero) to be kept.\
  If the user enters an invalid value, the script prints a friendly warning and asks again;
- keeps only OTUs whose prevalence is at least this threshold;
- then applies an **abundance filter**: among the remaining OTUs, it keeps only those whose median non-zero count is at least `abundance_threshold` (default = 5);
- prints a summary of the data **after** filtering and reports the thresholds used;
- returns (invisibly) a list containing:
  - `samp_filt`: the filtered OTU table;
  - `taxa_filt` (if a taxonomy table called `taxa` exists in the environment): the taxonomy table restricted to the surviving OTUs.

The function can be used both interactively (default, with prompt) and non-interactively (passing `prevalence_threshold` as an argument), which is useful for sensitivity analyses over a range of thresholds.

```{r}
# brings the filt_data script into scope
source(
  list.files(
    path = here(),
    pattern = "^filt_data\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


```

### pseudocount.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the pseudocount script into scope
source(
  list.files(
    path = here(),
    pattern = "^pseudocount\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


```

### generate_matrix_factors.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the generate_matrix_factors script into scope
source(
  list.files(
    path = here(),
    pattern = "^generate_matrix_factors\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

```

### pielou_ind.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the pielou_ind script into scope
source(
  list.files(
    path = here(),
    pattern = "^pielou_ind\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


```

## Orchestrators

<different clever text here>

### clr_pearson.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the clr_pearson script into scope
source(
  list.files(
    path = here(),
    pattern = "^clr_pearson\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


```

### NorTa_simulation.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the NorTa_simulation script into scope
source(
  list.files(
    path = here(),
    pattern = "^NorTa_simulation\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


```

### data_sim_ph_driven.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the data_sim_ph_driven script into scope
source(
  list.files(
    path = here(),
    pattern = "^data_sim_ph_driven\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


```

## Standalone analysis

Scripts in this section have the purpose of generating statistical analyses, each one evaluating a different issue that arose in the original paper.

### compositional_bias_D_P.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the compositional_bias_D_P script into scope
source(
  list.files(
    path = here(),
    pattern = "^compositional_bias_D_P\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


```

### ph_pielou_confrontation.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the ph_pielou_confrontation script into scope
source(
  list.files(
    path = here(),
    pattern = "^ph_dataset_diversity_confrontation\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


```
