---
title: "notebook_exam_statistical_data_analysis"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: false

# load important libraries
library(here)
library(tidyverse)
```

# Introduction

The aim of this notebook is to explain the scripts found in the folder `02_new_script/`.\
Scripts can be divided into three categories:

- **base functions**: small scripts that generate simple functions (a wise man once said: "if you have to do it more than once, instead write a script!").\
  The base functions are:

  - `datasummary.R`

  - `filt_data.R`

  - `pseudocount.R`

  - `generate_matrix_factors.R`

  - `pielou_ind.R`

- **orchestrators**: these are slightly more complex scripts than the base functions; their job is to use base functions to generate something more complex.\
  The orchestrators are:

  - `clr_pearson.R`

  - `NorTa_simulation.R`

  - `data_sim_ph_driven.R`

- **standalone analysis**: using the previous functions, these scripts generate the analyses that are important for this project. The standalone analyses are:

  - `compositional_bias_D_P.R`

  - `ph_pielou_confrontation.R`

Let's start by looking at the **base functions** scripts:

## Base functions

<clever text here>

### datasummary.R

*What it does:* defines the function `datasum()`, which prints a compact summary of any numeric dataset (tibble, data.frame, list, or matrix).\
The summary includes:

- dimensions (rows, columns);
- basic statistics: minimum, maximum, mean, median, standard deviation, skewness;
- zero structure: total number of zeroes and their percentage;
- for square numeric matrices only: determinant and minimum eigenvalue (useful to check properties of correlation/covariance matrices).

*Why is it here:* during the project I felt the need for a quick, consistent way to inspect matrices and tables (a wise man once said: "if you have to do it more than once, instead write a script about it").\
`datasum()` centralises these checks in a single reusable function, which is particularly handy for compositional and metagenomic data, where zeroes and matrix properties matter for statistical analyses.

*How it works:* the function `datasum()`:

- checks if `x` is a square numeric matrix and, if so and if there are no `NA`/infinite values, computes the determinant and minimum eigenvalue;

- coerces the input to a tibble (converting lists to data frames if needed);

- computes min, max, mean, median, standard deviation, and skewness (third standardized moment);

- counts zeroes and their percentage;

- prints all these quantities with a short label derived from the object name;

- returns them invisibly as a named list, so they can be reused programmatically.

```{r}
# brings the datasummary script into scope
source(
  list.files(
    path = here(),
    pattern = "^datasummary\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# generate a matrix that will be used as example
example_matrix <- matrix(c(22, 3, 5, 7, 9, 10, 0, 0, 1),
  nrow = 3,
  ncol = 3,
  dimnames = list(NULL, c("A", "B", "C"))
)

# print the matrix for confrontation
print(example_matrix)

# apply the datasum function from the scoped script
datasum(example_matrix)
```

### filt_data.R

*What it does:* filters a metagenomic OTU count table (samples × OTUs) by prevalence and abundance, and shows a before/after summary using `datasum()`.

*Why is it here:* metagenomic data (like HMP2 used for this project) are compositional and very sparse.\
Before any downstream analysis, it is important to remove rare and low-abundance OTUs that mostly add noise and can distort statistical results.

*How it works:* the script defines the function `filt_data()`, which:

- prints a summary of the data **before** filtering using `datasum()`;
- asks the user for a **prevalence threshold** (a number between 0 and 1), i.e. the minimum fraction of samples in which an OTU must be present (non-zero) to be kept.\
  If the user enters an invalid value, the script prints a friendly warning and asks again;
- keeps only OTUs whose prevalence is at least this threshold;
- then applies an **abundance filter**: among the remaining OTUs, it keeps only those whose median non-zero count is at least `abundance_threshold` (default = 5);
- prints a summary of the data **after** filtering and reports the thresholds used;
- returns (invisibly) a list containing:
  - `samp_filt`: the filtered OTU table;
  - `taxa_filt` (if a taxonomy table called `taxa` exists in the environment): the taxonomy table restricted to the surviving OTUs.

The function can be used both interactively (default, with prompt) and non-interactively (passing `prevalence_threshold` as an argument), which is useful for sensitivity analyses over a range of thresholds.

*Note:* due to the interactive nature of filt_data.R script, the following data has been generated with the script `for_notebook.R` and then loaded here.

```{r}
#| echo: false

# brings the filt_data script into scope
source(
  list.files(
    path = here(),
    pattern = "^filt_data\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


# -------- show non-filtered data --------

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# tell the user that the following are non-filtered data
cat("\nThe following data were generated witht the script 'data_sim_ph_driven.R':\n")

# brings into scope pre-generated data
not_filt_data <- readRDS(
  list.files(
    path = here(),
    pattern = "^data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)


# print the pre-generated data
print(not_filt_data)


# -------- show filtered data --------

# brings into scope pre-generated and then filtered data
filtered_data <- readRDS(
  list.files(
    path = here(),
    pattern = "^filtered_data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

cat("\nThese were the unfiltered data.\n")
cat("\nNow we will apply the filt_data() function with a threshold of 0.3, obtaining the
    folowwing data:\n")

# print the generated and filtered data
print(filtered_data)

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")


```

### pseudocount.R

*What it does:* defines the function `pseudocount()`, which converts count data to row-wise proportions and replaces zeroes with row-specific pseudocounts computed from a user-defined threshold.

*Why is it here:* before applying the Centered Log-Ratio (CLR) transformation, it is necessary to substitute the zeroes in the dataset with pseudocounts (log(0) gives −inf as a result, *kinda* difficult to use for statistical analysis).

*How it works:*

- the script first asks the user for a number between 0 and 1 (if the user does not provide a valid one, the script keeps asking for it), that will be used as threshold;

- it computes the detection limit as 1 / (the sum of the values in the row);

- the pseudocount is computed as the detection limit times the threshold provided by the user;

- it converts every row's counts into proportions;

- it substitutes every zero found in a row with the pseudocount for that row.

```{r}
#| echo: false

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# tell the user that the example matrix used before will be printed again
cat("\nNow we will print the example_matrix that we used before:\n")

# print the example matrix
print(example_matrix)

# tell the user that the matrix passed thorugh the pseudocount() function will be printed
cat("\nAfter using the pseudocount() function with 0.5 as threshold,
    the following matrix is obtained:\n")

# this matrix was computed outside the notebook and then hardcoded here to avoid
# to launch the pseudcocunt.R script that might have caused the chunk to freeze
pseudocounted_matrix <- matrix(
  c(0.759, 0.241, 0.0172, 0.25, 0.75, 0.0417, 0.312, 0.625, 0.0625),
  nrow = 3, ncol = 3,
  dimnames = list(NULL, c("A", "B", "C"))
)
# print the pseudocounted matrix
print(pseudocounted_matrix)

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

### generate_matrix_factors.R

*What it does:* defines the function `generate_matrix_factors()`, which builds a valid correlation matrix (symmetric, unit diagonal, positive semi-definite) with an exact, controlled zero pattern, by splitting taxa into `n_groups` latent groups.

*Why is it here:* in the data_sim_ph_driven.R script is used the function "rmvnorm"
("Random MultiVariateNORMal") which allows to generate random multivariate normal
distributions.  
One of the input for this function is the correlation matrix.  
Since I wanted to control as as possible the output of the data generation, I opted 
to write this script that allowed me to have a randomly generated correlation matrix.  

*How it works:* `generate_matrix_factors(n, n_groups, ...)`:

- randomly assigns each of the `n` taxa to one of `n_groups` groups;
- builds a loadings matrix $\Lambda$ where each taxon has a random value only in the column of its own group;
- computes $\Sigma = \Lambda\Lambda'$ (always PSD, whatever $\Lambda$ is);
- rescales it to a proper correlation matrix (unit diagonal);
- returns the matrix, the group assignment, and the exact number of zero pairs (taxa in different groups end up with correlation exactly 0).

```{r}
# brings the generate_matrix_factors script into scope
source(
  list.files(
    path = here(),
    pattern = "^generate_matrix_factors\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# generate a demostrative matrix with n = 10, n_groups = 2, seed = 42
corr_mat_notebook <- generate_matrix_factors(n = 10, n_groups = 3, seed = 42)

# print the actual matrix
print(corr_mat_notebook$mat)

# print a correlation matrix with no OTU correlated
not_corr_mat_notebook <- generate_matrix_factors(n = 10,n_groups = 10, seed = 42)

# print the non-correlation matrix
print(not_corr_mat_notebook$mat)


```

### pielou_ind.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the pielou_ind script into scope
source(
  list.files(
    path = here(),
    pattern = "^pielou_ind\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

## Orchestrators

<different clever text here>

### clr_pearson.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the clr_pearson script into scope
source(
  list.files(
    path = here(),
    pattern = "^clr_pearson\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

### NorTa_simulation.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the NorTa_simulation script into scope
source(
  list.files(
    path = here(),
    pattern = "^NorTa_simulation\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

### data_sim_ph_driven.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the data_sim_ph_driven script into scope
source(
  list.files(
    path = here(),
    pattern = "^data_sim_ph_driven\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

## Standalone analysis

Scripts in this section have the purpose of generating statistical analyses, each one evaluating a different issue that arose in the original paper.

### compositional_bias_D_P.R

*What it does:*

*Why is it here:*

*How it works:*

```{r}
# brings the compositional_bias_D_P script into scope
source(
  list.files(
    path = here(),
    pattern = "^compositional_bias_D_P\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

### ph_pielou_confrontation.R

*What it does:* uses `data_sim_ph_driven()` across a pH gradient, at five different levels of pH tolerance, and tracks how the Pielou index changes, showing that narrow-tolerance ("specialist") communities produce noisier, lower evenness, while wide-tolerance ("generalist") communities stay stable and high.

*Why is it here:* This is the final exame of this project. I needed something that could
test how data_sim_ph_driven.R could have been employed (in this case to simulate how
generalists vs specialists bacteria behave at the same pH).

*How it works:* loops over a pH gradient for five tolerance scenarios, computes `pielou_ind()` at each point, plots the five curves, and combines them into a single figure.

```{r}
# brings the ph_pielou_confrontation script into scope
source(
  list.files(
    path = here(),
    pattern = "^ph_dataset_diversity_confrontation\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```
