---
title: "notebook_exam_statistical_data_analysis"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Indice"
    toc-location: left
theme:
  light: flatly
  dark: darkly
editor: visual
---

```{r}
#| label: setup
#| include: false

# load important libraries
library(here)
library(tidyverse)
```

# Introduction {style="color:orange"}

The aim of this notebook is to explain the scripts found in the folder `02_new_scripts/`.\
Some scripts require the user to provide an input so, instead of sourcing and running them in this notebook, some results have been generated outside this script (using the script `for_notebook.R`) and then pasted in their relative code chunks.\
To separate the code chunks from the plain text, spacing bars like this

```{r}
#| echo: false
# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

were put at the begin and at the end of the code chunks that they need to separe.\
Scripts can be divided into three categories:

- [**base functions**]{style="color:violet"}: small scripts that generate simple functions (a wise man once said: "if you have to do it more than once, instead write a script!").\
  The base functions are:

  - `datasummary.R`

  - `filt_data.R`

  - `pseudocount.R`

  - `generate_matrix_factors.R`

  - `pielou_ind.R`

- [**orchestrators**]{style="color:violet"}: these are slightly more complex scripts than the base functions; their job is to use base functions to generate something more complex.\
  The orchestrators are:

  - `clr_pearson.R`

  - `NorTa_simulation.R`

  - `data_sim_ph_driven.R`

- [**standalone analysis**]{style="color:violet"}: using the previous functions, these scripts generate the analyses that are important for this project. The standalone analyses are:

  - `compositional_bias_D_P.R`

  - `ph_dataset_diversity_confrontation.R`

Let's start by looking at the **base functions** scripts:

## [Base functions]{style="color:violet"}

Base functions are the smallest building blocks of this project: each one does a single, well-defined job (summarizing data, filtering, replacing zeroes, generating a correlation matrix, or computing diversity), and gets reused by the more complex scripts (orchestrators and standalone analysis) described later.

### [datasummary.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} defines the function `datasum()`, which prints a compact summary of any numeric dataset (tibble, data.frame, list, or matrix).\
The summary includes:

- dimensions (rows, columns);
- basic statistics: minimum, maximum, mean, median, standard deviation, skewness;
- zero structure: total number of zeroes and their percentage;
- for square numeric matrices only: determinant and minimum eigenvalue (useful to check properties of correlation/covariance matrices).

[*Why is it here:*]{style="color:lime"} during the project, a quick and consistent way to inspect matrices and tables was needed.\
`datasum()` centralises these checks in a single reusable function, which is particularly handy for compositional and metagenomic data, where zeroes and matrix properties matter for statistical analyses.

[*How it works:*]{style="color:lime"} the function `datasum()`:

- checks if `x` is a square numeric matrix and, if so and if there are no `NA`/infinite values, computes the determinant and minimum eigenvalue;

- coerces the input to a tibble (converting lists to data frames if needed);

- computes min, max, mean, median, standard deviation, and skewness (third standardized moment);

- counts zeroes and their percentage;

- prints all these quantities with a short label derived from the object name;

- returns them invisibly as a named list, so they can be reused programmatically.

```{r}
# brings the datasummary script into scope
source(
  list.files(
    path = here(),
    pattern = "^datasummary\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# generate an example matrix
example_matrix <- matrix(c(22, 3, 5, 7, 9, 10, 0, 0, 1),
  nrow = 3,
  ncol = 3,
  dimnames = list(c("alfa", "beta", "gamma"), c("A", "B", "C"))
)

# print the example matrix
print(example_matrix)

# apply the datasum function from the scoped script
datasum(example_matrix)
```

### [filt_data.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} filters a metagenomic OTU count table (samples × OTUs) by prevalence and abundance, and shows a before/after summary using `datasum()`.

[*Why is it here:*]{style="color:lime"} metagenomic data (like HMP2 used in the original paper and for this project) are compositional and very sparse.\
Before any downstream analysis, it is important to remove rare and low-abundance OTUs that mostly add noise and can distort statistical results.

[*How it works:*]{style="color:lime"} the script defines the function `filt_data()`, which:

- prints a summary of the data **before** filtering using `datasum()`;
- asks the user for a **prevalence threshold** (a number between 0 and 1), i.e. the minimum fraction of samples in which an OTU must be present (non-zero) to be kept.\
  If the user enters an invalid value, the script prints a friendly warning and asks again;
- keeps only OTUs whose prevalence is at least this threshold;
- then applies an **abundance filter**: among the remaining OTUs, it keeps only those whose median non-zero count is at least `abundance_threshold` (default = 5);
- prints a summary of the data **after** filtering and reports the thresholds used;
- returns (invisibly) a list containing:
  - `samp_filt`: the filtered OTU table;
  - `taxa_filt` (if a taxonomy table called `taxa` exists in the environment): the taxonomy table restricted to the surviving OTUs.

The function can be used both interactively (default, with prompt) and non-interactively (passing `prevalence_threshold` as an argument), which is useful for sensitivity analyses over a range of thresholds.

[*Note:* due to the interactive nature of the filt_data.R script (the script asks the user for a number between 0 and 1 which is used as threshold to set the filter), the following data has been generated with the script `for_notebook.R`, then loaded here.]{style="color:red"}

```{r}
#| echo: false

# brings the filt_data script into scope
source(
  list.files(
    path = here(),
    pattern = "^filt_data\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


# -------- show non-filtered data --------

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# tell the user that the following are non-filtered data
cat("\nThe following data were generated with the script 'data_sim_ph_driven.R':\n")

# brings into scope pre-generated data
not_filt_data <- readRDS(
  list.files(
    path = here(),
    pattern = "^data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)


# print the pre-generated data
print(not_filt_data)


# -------- show filtered data --------

# brings into scope pre-generated and then filtered data
filtered_data <- readRDS(
  list.files(
    path = here(),
    pattern = "^filtered_data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

cat("\nThese were the unfiltered data.\n")
cat("\nNow we will apply the filt_data() function with a threshold of 0.3, obtaining the
    following data:\n")

# print the generated and filtered data
print(filtered_data)

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

### [pseudocount.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} defines the function `pseudocount()`, which converts count data to row-wise proportions and replaces zeroes with row-specific pseudocounts computed from a user-defined threshold.

[*Why is it here:*]{style="color:lime"} before applying the Centered Log-Ratio (CLR) transformation, the zeroes in the dataset need to be replaced with pseudocounts, since `log(0)` returns `-Inf`, which is not usable in statistical analysis.

[*How it works:*]{style="color:lime"}

- the script first asks the user for a number between 0 and 1 (if the user does not provide a valid one, the script keeps asking for it), that will be used as threshold;

- it computes the detection limit as 1 / (the sum of the values in the row);

- the pseudocount is computed as the detection limit times the threshold provided by the user;

- it converts every row's counts into proportions;

- it substitutes every zero found in a row with the pseudocount for that row.

[*Note:* due to the interactive nature of the pseudocount.R script (the script asks the user for a number between 0 and 1 which is used as threshold to set the filter), the following data has been generated with the script `for_notebook.R`, then loaded here.]{style="color:red"}

```{r}
#| echo: false

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# tell the user that the example matrix used before will be printed again
cat("\nNow we will print the example_matrix that we used before:\n")

# print the example matrix
print(example_matrix)

# tell the user that the matrix passed through the pseudocount() function will be printed
cat("\nAfter using the pseudocount() function with 0.5 as threshold,
    the following matrix is obtained:\n")

# this matrix was computed outside the notebook and then hardcoded here to
# avoid launching the pseudocount.R script, which might cause the chunk to freeze
pseudocounted_matrix <- matrix(
  c(0.759, 0.241, 0.0172, 0.25, 0.75, 0.0417, 0.312, 0.625, 0.0625),
  nrow = 3, ncol = 3,
  dimnames = list(c("alfa", "beta", "gamma"), c("A", "B", "C"))
)
# print the pseudocounted matrix
print(round(pseudocounted_matrix, 2))

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

### [generate_matrix_factors.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} defines the function `generate_matrix_factors()`, which builds a valid `nxn` correlation matrix (symmetric, unit diagonal, positive semi-definite, values from -1 to +1) with an exact, controlled zero pattern, by splitting taxa into `n_groups` latent groups.

[*Why is it here:*]{style="color:lime"} in the data_sim_ph_driven.R script, the function "rmvnorm" ("Random MultiVariateNORMal") is used, which makes it possible to generate random multivariate normal distributions.\
One of the inputs of the "rmvnorm" function is the correlation matrix. Using this script makes it possible to obtain a controlled (and yet partially randomized) correlation matrix.

[*How it works:*]{style="color:lime"} `generate_matrix_factors(n, n_groups, ...)` builds the correlation matrix indirectly, through a small latent-factor model, instead of generating random values directly and then fixing them up to be a valid correlation matrix:

- randomly assigns each of the `n` taxa to one of `n_groups` groups;

- builds a loadings matrix $\Lambda$ (dimension `n x n_groups`): each taxon (row) gets a single random value only in the column of its own group, and 0 everywhere else. In other words, each taxon's loading vector "points" only along its own group's axis;

[*Note*: so far we have seen taxa placed only in columns, while in this specific passage they are put in rows. This was necessary due to how the matrix product works since $\Lambda$ = `n x n_groups` then $\Lambda'$ must be = `n_groups x n`. This is necessary to obtain a final `nxn` matrix]{style="color:red"}

- computes $\Sigma = \Lambda\Lambda'$: this is the **Gram matrix** of the taxa's loading vectors, i.e. the matrix of all pairwise dot products between them. Two taxa in **different** groups have loading vectors that are non-zero in different columns, so their dot product is exactly 0 (no shared direction); two taxa in the **same** group have a dot product equal to the product of their two loading values (generally non-zero). This is exactly why the zero pattern of $\Sigma$ is controlled and exact, not just approximately zero;

- because $\Sigma$ is built as $\Lambda\Lambda'$, it is guaranteed to be positive semi-definite (PSD) for *any* $\Lambda$: for any vector $x$, $x'\Sigma x = x'\Lambda\Lambda'x = \lVert \Lambda' x \rVert^2 \geq 0$. This is the whole point of using a Gram matrix construction here: it sidesteps the usual problem of generating a random matrix and then having to "fix" it into a valid (PSD) correlation matrix afterwards (e.g. via `nearPD()`), at the cost of only being able to reach correlation structures that come from this latent-group form;

- rescales $\Sigma$ to a proper correlation matrix (dividing by the geometric mean of each pair's standard deviations, so the diagonal becomes exactly 1);

- returns the matrix, the group assignment, and the exact number of zero pairs (taxa in different groups end up with correlation exactly 0).

```{r}
# brings the generate_matrix_factors script into scope
source(
  list.files(
    path = here(),
    pattern = "^generate_matrix_factors\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# generate a demonstrative matrix with taxa = n = 10, number of groups = n_groups = 3, seed = 42
corr_mat_notebook <- generate_matrix_factors(n = 10, n_groups = 3, seed = 42)

# print the actual matrix
print(corr_mat_notebook$mat)

# print a correlation matrix with no OTU correlated
not_corr_mat_notebook <- generate_matrix_factors(n = 10, n_groups = 10, seed = 42)

# print the non-correlation matrix
print(not_corr_mat_notebook$mat)
```

### [pielou_ind.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} defines the function `pielou_ind()`, which computes the dataset diversity as the mean of the pielou indexes of all the samples in the dataset.

[*Why is it here:*]{style="color:lime"} it is a good parameter (also used in the original paper) that makes it possible to quantify the species diversity of a sample.\
It is needed for the final analysis.

[*How it works:*]{style="color:lime"} `pielou_ind(x, per_sample = FALSE)`:

- computes each sample's relative abundances and Shannon entropy, handling zeroes with the standard convention $0 \cdot \ln(0) = 0$, so it works directly on sparse real data, no pseudocounts needed;
- normalizes by $\ln(D)$ (D = number of taxa), so the index stays comparable across datasets of different size;
- returns the mean across samples by default, or the full per-sample vector if `per_sample = TRUE`.

```{r}
#| echo: false
# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")


# -------- sourced stuff needed for this chunk --------

# brings the pielou_ind script into scope
source(
  list.files(
    path = here(),
    pattern = "^pielou_ind\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# read previously generated data
data_for_notebook_count <- readRDS(
  list.files(
    path = here(),
    pattern = "^data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# -------- display of script activity --------

# shows the sourced data
print(data_for_notebook_count)

# explain the user what's happening
cat("\nApplying the pielou_ind() function to the dataset
(using per_sample = FALSE),
we obtain the within dataset diversity parameter:\n")

# apply the pielou_ind() function to the sourced data (per_sample = FALSE will
# give the within dataset diversity parameter):
round(pielou_ind(data_for_notebook_count, per_sample = FALSE), 2)

# explain the user what's happening
cat("\nNow, if we use the same function with per_sample = TRUE,
we obtain a vector with all of the pielou indexes of the
data (data has 8 samples, which means 8 rows, so we will
obtain 8 indexes):\n")

# apply the pielou_ind() function to the sourced data using per_sample = TRUE
pi_ind_vect <- pielou_ind(data_for_notebook_count, per_sample = TRUE)

# show the various pielou indexes computed from the dummy dataset
print(pi_ind_vect, digits = 2)

# explain the user what's happening
cat("\nIf we compute the mean of the previously generated vector,
we obtain again the within dataset diversity parameter:\n")

# show that applying the mean to the pi_ind_vect gives us the already printed within
# dataset diversity parameter
print(round(mean(pi_ind_vect), 2))

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

## [Orchestrators]{style="color:violet"}

Orchestrators are the next level of base functions: instead of doing a single small job, they combine several base functions together into a full step of the analysis (filtering + pseudocount + CLR + correlation, or building a correlation matrix + simulating data from it).

### [clr_pearson.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} defines the function `clr_on_data()`, which runs the full pipeline on an OTU table: - filters it (`filt_data()`), - replaces zeroes with pseudocounts (`pseudocount()`), - applies the CLR transformation, - computes the OTU x OTU Pearson correlation matrix.

[*Why is it here:*]{style="color:lime"} one of the main takeaways of the original paper is that the "blessing of dimensionality" makes the use of the centered log-ratio transformation in combination with Pearson's correlation coefficient (from now on, clr+Pearson) a reliable tool for analyzing correlations between taxa. Moreover, clr+Pearson is used as a benchmark to evaluate how other methods behave in a high-dimensional setting (dimensionality = D = number of taxa \> 100). Given the importance of this aspect, the method was replicated here, with the added possibility of choosing which filters to apply to the data.

[*How it works:*]{style="color:lime"} `clr_on_data(x)` runs the four steps above in sequence and returns (invisibly) a list with `samp_filt` (the filtered table), `y_clr` (the CLR-transformed table), and `cor_matrix` (the resulting correlation matrix).

[*Note:* due to the interactive nature of the clr_pearson.R script (it uses both filt_data() and pseudocount(), which both ask for user input), the source() call below is commented out, and the following data has been generated with the script `for_notebook.R`, then loaded here.]{style="color:red"}

```{r}
#| echo: false

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# brings the clr_pearson script into scope
# source(
#   list.files(
#     path = here(),
#     pattern = "^clr_pearson\\.R$",
#     full.names = TRUE,
#     recursive = TRUE
#   )
# )

# brings the pre-generated dummy data into scope
data_for_notebook_count <- readRDS(
  list.files(
    path = here(),
    pattern = "^data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# cat to tell the user that we will re-print some data
cat("\nNow we will re-print some data that we already used:\n")
# print the pre-generated data
print(data_for_notebook_count)

# tell the user the parameters used in the clr_on_data() processing
cat("\nWe will now apply the clr_on_data() function to the previous data, 
using 0.33 as prevalence threshold for the filt_data() 
part and 0.5 for the pseudocount() part.\n")

# introduce the user to the results
cat("\nThe results of the pipeline are the following:\n")


# -------- brings the three separate elements of the list generated by the clr_on_data() function and print them --------

# brings the filtered sample into scope
processed_data_samp_filt <- readRDS(
  list.files(
    path = here(),
    pattern = "^processed_data_samp_filt\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# brings the log-transformed data into scope
processed_data_y_clr <- readRDS(
  list.files(
    path = here(),
    pattern = "^processed_data_y_clr\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# brings the correlation matrix of the transformed data into scope
processed_data_cor_mat <- readRDS(
  list.files(
    path = here(),
    pattern = "^processed_data_cor_matrix\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# title of the filtered sample
cat("\nFiltered sample\n")
# prints the filtered sample
print(processed_data_samp_filt)

# title of the log transformed data
cat("\nLog transformed data\n")
# prints the log transformed data
print(round(processed_data_y_clr, 2))

# title of the printed correlation matrix
cat("\nCorrelation matrix\n")
# print the correlation matrix
print(processed_data_cor_mat)

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

```

### [NorTa_simulation.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} defines the function `norta_simulation()`, which builds a correlation matrix with `generate_matrix_factors()` and then simulates a sparse count dataset from it, following the NorTA ("Normal To Anything") approach: correlated normal data are converted to uniform percentiles and then mapped through the quantile function of a Zero-Inflated Negative Binomial (ZINB) distribution.

[*Why is it here:*]{style="color:lime"} the genesis of this script was part of the work that lead me to obtain `data_sim_ph_driven.R`.\
I needed a script that allowed me to generate compositional data.\
Why though? Because to validate a correlation-estimation method, it is useful to have simulated data with a known correlation structure.\
This was part of the work in the original paper and I wanted to try to create my version for it.

[*How it works:*]{style="color:lime"} `norta_simulation(n, n_groups, N, mu, size, phi, seed)`:

- builds the correlation matrix and simulates `N` correlated normal samples from it;
- converts the normal data to uniform percentiles via `pnorm()` (a monotonic transformation, so rank correlation is preserved);
- maps the percentiles through the ZINB quantile function, using a single `phi` (zero-inflation probability) shared by all taxa.

```{r}
#| echo: false

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# brings into scope the NorTa_simulation.R script
source(
  list.files(
    path = here(),
    pattern = "^NorTa_simulation\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# explain the user how the datas are generated with which parameters
cat("\nThe following data are generated with the norta_simulation()
function defined in the NorTa_simulation.R script, using the
following parameters:
      - n = 9;\n
      - n_groups = 3;\n
      - N = 11;\n
      - seed = 42;\n")

# additional space 
cat("\n")

# generate data with n = 9, n_groups = 3, N = 11, seed = 42
norta_data_01 <- norta_simulation(n = 9, n_groups = 3, N = 11, seed = 42) 

# title of the data
cat("\nData generated with n_groups = 3 and n = 9\n")
# print generated data sim_counts
print(norta_data_01$sim_counts)

# title of the first correlation matrix
cat("\nCorrelation matrix with n_groups = 3 and n = 9\n")
# print generated data sim_counts
print(norta_data_01$R_true)

# explain the user how the datas are generated with which parameters
cat("\nThe following data are generated with the norta_simulation()
function defined in the NorTa_simulation.R script, using the
following parameters:
      - n = 10;\n
      - n_groups = 10;\n
      - N = 10;\n
      - seed = 42;\n
With a number of groups (n_groups) equal to the number of taxa 
(number of taxa is the parameter n), it will resolve in having
zero taxa correlated to each other, leading to a unitary correlation matrix.\n")

# additional space 
cat("\n")

# generate data with n = 10, n_groups = 10, N = 10, seed = 42 (gives an identity matrix)
norta_data_02 <- norta_simulation(n = 10, n_groups = 10, N = 10, seed = 42) 

# title of the data
cat("\nData generated with n_groups = n\n")
# print generated data sim_counts
print(norta_data_02$sim_counts)

# title of the second correlation matrix
cat("\nCorrelation matrix with n_groups = n\n")
# print generated data sim_counts
print(norta_data_02$R_true)


# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

```

### [data_sim_ph_driven.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} follows the same NorTA approach as `NorTa_simulation.R`, but sparsity is no longer a single value shared by all taxa: each taxon is given its own zero-inflation probability, based on the distance between the environmental pH and that taxon's own pH optimum.

[*Why is it here:*]{style="color:lime"} this script represents the main extension proposed in this project.\
Rather than treating sparsity as an arbitrary simulation parameter, it derives it from a biological ambiental parameter (pH), so that the resulting sparsity pattern has an ecological interpretation.

[*How it works:*]{style="color:lime"} `data_sim_ph_driven(n_taxa, N_sample, ph, n_groups, ...)`:

- assigns each taxon a random pH optimum and tolerance (niche width);
- computes each taxon's zero-inflation probability from a Gaussian "suitability" curve centred on its own optimum: well-matched taxa get a low zero-inflation probability, poorly-matched taxa get a high one;
- runs the same NorTA steps as `NorTa_simulation.R`, using this per-taxon zero-inflation vector instead of a single shared value.

Unlike `filt_data()`, `pseudocount()` and `clr_on_data()`, this function has **no interactive input**: every parameter (including `ph`, `sigma_min`/`sigma_max` and `phi_max`) is an explicit argument, so it can be sourced and run directly in this notebook, exactly like `NorTa_simulation.R`.

To make the pH-driven mechanism visible, the same set of taxa (same `seed`, same pH optima/tolerances) is simulated twice, changing only the community's environmental `ph`: once close to the taxa's average optimum, and once shifted towards a more acidic environment. Comparing `phi_per_taxon` and `total_zero_rate` between the two runs shows the mechanism at work: taxa whose optimum is far from the community pH become sparser, without ever touching a fixed, shared zero-inflation parameter.

```{r}
#| echo: false

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# brings the data_sim_ph_driven script into scope
source(
  list.files(
    path = here(),
    pattern = "^data_sim_ph_driven\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# explain the user how the data are generated and with which parameters
cat("\nThe following data are generated with the data_sim_ph_driven()
function defined in the data_sim_ph_driven.R script, using the
following parameters:
      - n_taxa = 6;\n
      - N_sample = 8;\n
      - n_groups = 2;\n
      - ph = 6.5 (close to the middle of the default niche range);\n
      - seed = 42;\n")

# additional space
cat("\n")

# generate the dataset for a community with pH = 6.5
sim_ph_mid <- data_sim_ph_driven(
  n_taxa = 6, N_sample = 8, ph = 6.5, n_groups = 2, seed = 42
)

# title of the per-taxon pH optima
cat("\npH optimum assigned to each taxon (same for both simulations below,
since the same seed is used):\n")
# print the pH optimum of every taxon
print(round(sim_ph_mid$ph_optima, 2))

# title of the resulting zero-inflation probabilities
cat("\nZero-inflation probability per taxon, with community pH = 6.5:\n")
# print the zero-inflation probability of every taxon
print(round(sim_ph_mid$phi_per_taxon, 2))

# title of the overall zero rate
cat("\nOverall fraction of zeroes in the simulated counts (pH = 6.5):\n")
# print the achieved overall zero rate
print(round(sim_ph_mid$total_zero_rate, 2))

# explain what is changing in the second simulation
cat("\nWe now repeat the exact same simulation (same seed, same taxa,
same pH optima and tolerances), changing only the community pH to 5.6
(a more acidic environment):\n")

# additional space
cat("\n")

# generate the dataset for a more acidic community, pH = 5.6
sim_ph_acidic <- data_sim_ph_driven(
  n_taxa = 6, N_sample = 8, ph = 5.6, n_groups = 2, seed = 42
)

# title of the resulting zero-inflation probabilities
cat("\nZero-inflation probability per taxon, with community pH = 5.6:\n")
# print the zero-inflation probability of every taxon
print(round(sim_ph_acidic$phi_per_taxon, 2))

# title of the overall zero rate
cat("\nOverall fraction of zeroes in the simulated counts (pH = 5.6):\n")
# print the achieved overall zero rate
print(round(sim_ph_acidic$total_zero_rate, 2))

# explain the comparison between the two simulations
cat("\nComparing the two phi_per_taxon vectors, taxa with a pH optimum
close to 5.6 now have a lower zero-inflation probability than before,
while taxa optimized for higher pH values become sparser: the same
taxon can go from 'common' to 'rare' purely because the environmental
pH changed, without touching any shared/fixed sparsity parameter.\n")

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

## [Standalone analysis]{style="color:violet"}

The final level of complexity of all the project.\
The scripts presented in this section perform the main statistical analyses of the project.\
Each analysis is designed to investigate a specific issue identified in the original paper and to assess its potential implications for the results.

### [compositional_bias_D_P.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} replicates Section 3.1 of the original paper: it quantifies how much the L1 and CLR normalizations distort correlation estimates, as dimensionality (D) and within-dataset diversity (P) vary.\
The ground truth used is a fully uncorrelated dataset (true correlation exactly zero everywhere), so that any non-zero correlation observed after normalization can be attributed entirely to the normalization method itself.

[*Why is it here:*]{style="color:lime"} it provides a direct, quantitative answer to whether CLR actually outperforms L1, and by how much, depending on dimensionality and diversity, confirming the results obtained in the original paper.  
This script exists, basically, because I wanted to replicate the paper's results myself, rather than taking them at face value. 

[*How it works:*]{style="color:lime"} for each (D, P) combination on a 15x15 grid (a reduced version of the paper's 40x39 grid, using N = 5,000 samples instead of 10,000, for time constraints):

- generates uncorrelated Gaussian data and tunes its diversity to the target P (via `pielou_ind()` and a numerical root search);
- applies the L1 and CLR normalizations and computes the resulting correlation matrices;
- records the mean absolute error of each method against the true (zero) correlation.

```{r}
# brings the compositional_bias_D_P script into scope
source(
  list.files(
    path = here(),
    pattern = "^compositional_bias_D_P\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# display the heatmap previously generated by compositional_bias_D_P.R
knitr::include_graphics(
  list.files(
    path = here(),
    pattern = "^compositional_bias_heatmap\\.png$",
    full.names = TRUE,
    recursive = TRUE
  )
)

```

### [ph_dataset_diversity_confrontation.R]{style="color:cyan"}

[*What it does:*]{style="color:lime"} uses `data_sim_ph_driven()` across a pH gradient, at five different levels of pH tolerance, and tracks how the within dataset diversity parameter changes, showing that narrow-tolerance ("specialist") communities produce noisier, lower evenness, while wide-tolerance ("generalist") communities stay stable and high.

[*Why is it here:*]{style="color:lime"} this is the final script presented in this project. `data_sim_ph_driven.R` is, in my humble opinion, a good simulation tool, but a tool with no job to do is just a tool, so it needed to be put to work and see what it could actually produce.  
Rather than reaching for something overly elaborate, I chose to explore a result most people would already expect intuitively: specialist taxa tend to generate environments with lower diversity.

[*How it works:*]{style="color:lime"} loops over a pH gradient for five tolerance scenarios, computes `pielou_ind()` at each point, plots the five curves, and combines them into a single figure.

```{r}
# brings the ph_dataset_diversity_confrontation script into scope
source(
  list.files(
    path = here(),
    pattern = "^ph_dataset_diversity_confrontation\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

## [Project explanation]{style="color:violet"}

The previous sections documented each script *in isolation*: what it does, why it exists, and how it works, each with a small, self-contained example. This section instead follows the *actual logical path* of the project, from the initial research question to the final results, showing how the base functions and orchestrators are chained together into the two standalone analyses.  
In short: earlier, we looked at the tools in the toolbox, now we look why and how
we needed those tools.

### [Starting point and research question]{style="color:cyan"}

The project starts from a specific claim made in the original paper (Fuschi et al., 2025): that the CLR transformation, combined with Pearson's correlation, is a reliable way to estimate taxon-taxon correlations in high-dimensional, compositional metagenomic data, and that this reliability improves with dimensionality (the "blessing of dimensionality").  
Two questions follow directly from this claim:

1.  *How much* compositional bias does CLR actually remove compared to a simpler normalization (L1), and how does this depend on the number of taxa (D) and on how evenly abundances are distributed within a sample (P, within dataset diversity)?
2.  Can the sparsity patterns typically simulated to test this kind of method be made more biologically realistic, instead of relying on a single, arbitrary zero-inflation parameter shared by every taxon?

These two questions map directly onto the two standalone analyses of the project.

### [Building the tools, in the order they were needed]{style="color:cyan"}

To answer question 1 (`compositional_bias_D_P.R`), three ingredients were needed first:

- a way to measure within-dataset diversity for any simulated dataset → `pielou_ind()`;
- a way to compute CLR (reused directly inside `compositional_bias_D_P.R`, following the same logic already implemented for real data in `clr_pearson.R`);
- no correlation matrix generator was needed here, since the ground truth for this analysis is deliberately the identity matrix (zero true correlation everywhere), so that any correlation measured after normalization is attributable only to the compositional constraint.

To answer question 2 (`data_sim_ph_driven.R`, used in `ph_dataset_diversity_confrontation.R`), a different chain of tools was needed:

- a way to generate a controlled, valid correlation matrix for the simulated taxa → `generate_matrix_factors()`;
- a way to turn that correlation matrix into a realistic, sparse count dataset → the NorTA approach, first implemented with a single shared zero-inflation parameter in `NorTa_simulation.R`;
- the actual extension of the project: replacing that single shared parameter with a per-taxon, pH-driven zero-inflation probability → `data_sim_ph_driven.R`;
- `pielou_ind()` again, this time to measure how the *pattern* of sparsity (not just its overall amount) affects within-dataset diversity along a pH gradient.

In parallel, on the real-data side of the project, `datasummary.R`, `filt_data.R` and `pseudocount.R` were written to bring a real OTU table into a state where CLR is applicable (no zeroes, low-quality OTUs removed), and `clr_pearson.R` orchestrates them into the same CLR + Pearson pipeline that the simulation-based analyses are meant to stress-test.

### [The full pipeline]{style="color:cyan"}

Putting everything together, the analysis follows this order:

1.  **Real-data pipeline** (`clr_pearson.R`, using `datasummary.R`, `filt_data.R`, `pseudocount.R`): a raw OTU table is summarized, filtered, pseudocounted and CLR-transformed, producing an OTU x OTU Pearson correlation matrix. This is the concrete instance of the method whose reliability the original paper discusses, and which the rest of the project investigates more systematically.

2.  **Compositional bias analysis** (`compositional_bias_D_P.R`, using `pielou_ind()`): stepping back from any single real dataset, this analysis asks how much bias CLR (and, for comparison, L1) introduces *in general*, as a function of D (dimensionality) and P (within dataset diversity), using fully uncorrelated synthetic data as ground truth.  
The output (the MAE heatmap) is a direct, quantitative confirm on the "blessing of dimensionality" claim.

3.  **Realistic sparsity simulation** (`generate_matrix_factors.R` → `NorTa_simulation.R` → `data_sim_ph_driven.R`): moving from an arbitrary, single zero-inflation parameter to a per-taxon, ecologically motivated one, this chain produces simulated datasets whose sparsity pattern has a biological interpretation (pH niches), rather than being a free simulation knob.

4.  **pH-driven diversity analysis** (`ph_dataset_diversity_confrontation.R`, using `data_sim_ph_driven()` and `pielou_ind()`): the pH-driven simulator is finally used to study how community evenness responds to an environmental gradient, contrasting specialist (narrow pH tolerance) and generalist (wide pH tolerance) taxa.

Steps 1-2 and 3-4 can be run independently of each other: they answer the two separate research questions above, and neither depends on the other's output. What they share is the same underlying toolkit (`pielou_ind()`, and the CLR logic), which is why both are presented in the same project rather than as two unrelated scripts.

### [How to reproduce it]{style="color:cyan"}

To reproduce the whole analysis from scratch (as opposed to the pre-generated `.rds` files loaded above for the interactive scripts):

- run `compositional_bias_D_P.R` on its own: it has no interactive input and, when sourced, regenerates `outputs/compositional_bias_results.rds` and `outputs/compositional_bias_heatmap.png` from scratch;
- run `ph_dataset_diversity_confrontation.R` on its own: it also has no interactive input, and regenerates `Plots/ph_dataset_diversity_confrontation.png` from scratch;
- for the real-data pipeline (`clr_pearson.R`), either supply `prevalence_threshold` explicitly to `filt_data()` and answer the `pseudocount()` prompt, or use `for_notebook.R` as a template for a non-interactive run with fixed thresholds, exactly as was done to produce the `.rds` files loaded in this notebook.

## [Conclusions]{style="color:cyan"}
