---
title: "notebook_exam_statistical_data_analysis"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: false

# load important libraries
library(here)
library(tidyverse)
```

# Introduction

The aim of this notebook is to explain the scripts found in the folder `02_new_script/`.\
Scripts can be divided into three categories:

- **base functions**: small scripts that generate simple functions (a wise man once said: "if you have to do it more than once, instead write a script!").\
  The base functions are:

  - `datasummary.R`

  - `filt_data.R`

  - `pseudocount.R`

  - `generate_matrix_factors.R`

  - `pielou_ind.R`

- **orchestrators**: these are slightly more complex scripts than the base functions; their job is to use base functions to generate something more complex.\
  The orchestrators are:

  - `clr_pearson.R`

  - `NorTa_simulation.R`

  - `data_sim_ph_driven.R`

- **standalone analysis**: using the previous functions, these scripts generate the analyses that are important for this project. The standalone analyses are:

  - `compositional_bias_D_P.R`

  - `ph_pielou_confrontation.R`

Let's start by looking at the **base functions** scripts:

## Base functions

Base functions are the smallest building blocks of this project: each one does a single, well-defined job (summarizing data, filtering, replacing zeroes, generating a correlation matrix, or computing diversity), and gets reused by the more complex scripts (orchestrators and standalone analysis) described later.


### datasummary.R

[*What it does:*]{style="color:green"} defines the function `datasum()`, which prints a compact summary of any numeric dataset (tibble, data.frame, list, or matrix).\
The summary includes:

- dimensions (rows, columns);
- basic statistics: minimum, maximum, mean, median, standard deviation, skewness;
- zero structure: total number of zeroes and their percentage;
- for square numeric matrices only: determinant and minimum eigenvalue (useful to check properties of correlation/covariance matrices).

[*Why is it here:*]{style="color:green"} during the project I felt the need for a quick, consistent way to inspect matrices and tables.\
`datasum()` centralises these checks in a single reusable function, which is particularly handy for compositional and metagenomic data, where zeroes and matrix properties matter for statistical analyses.

[*How it works:*]{style="color:green"} the function `datasum()`:

- checks if `x` is a square numeric matrix and, if so and if there are no `NA`/infinite values, computes the determinant and minimum eigenvalue;

- coerces the input to a tibble (converting lists to data frames if needed);

- computes min, max, mean, median, standard deviation, and skewness (third standardized moment);

- counts zeroes and their percentage;

- prints all these quantities with a short label derived from the object name;

- returns them invisibly as a named list, so they can be reused programmatically.

```{r}
# brings the datasummary script into scope
source(
  list.files(
    path = here(),
    pattern = "^datasummary\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# generate an example matrix
example_matrix <- matrix(c(22, 3, 5, 7, 9, 10, 0, 0, 1),
  nrow = 3,
  ncol = 3,
  dimnames = list(c("alfa", "beta", "gamma"), c("A", "B", "C"))
)

# print the example matrix
print(example_matrix)

# apply the datasum function from the scoped script
datasum(example_matrix)
```

### filt_data.R

[*What it does:*]{style="color:green"} filters a metagenomic OTU count table (samples × OTUs) by prevalence and abundance, and shows a before/after summary using `datasum()`.

[*Why is it here:*]{style="color:green"} metagenomic data (like HMP2 used for this project) are compositional and very sparse.\
Before any downstream analysis, it is important to remove rare and low-abundance OTUs that mostly add noise and can distort statistical results.

[*How it works:*]{style="color:green"} the script defines the function `filt_data()`, which:

- prints a summary of the data **before** filtering using `datasum()`;
- asks the user for a **prevalence threshold** (a number between 0 and 1), i.e. the minimum fraction of samples in which an OTU must be present (non-zero) to be kept.\
  If the user enters an invalid value, the script prints a friendly warning and asks again;
- keeps only OTUs whose prevalence is at least this threshold;
- then applies an **abundance filter**: among the remaining OTUs, it keeps only those whose median non-zero count is at least `abundance_threshold` (default = 5);
- prints a summary of the data **after** filtering and reports the thresholds used;
- returns (invisibly) a list containing:
  - `samp_filt`: the filtered OTU table;
  - `taxa_filt` (if a taxonomy table called `taxa` exists in the environment): the taxonomy table restricted to the surviving OTUs.

The function can be used both interactively (default, with prompt) and non-interactively (passing `prevalence_threshold` as an argument), which is useful for sensitivity analyses over a range of thresholds.

[*Note:* due to the interactive nature of the filt_data.R script (the script asks the user for a number between 0 and 1 which is used as threshold to set the filter) , the following data has been generated with the script `for_notebook.R`, then loaded here.]{style="color:red"}

```{r}
#| echo: false

# brings the filt_data script into scope
source(
  list.files(
    path = here(),
    pattern = "^filt_data\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)


# -------- show non-filtered data --------

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# tell the user that the following are non-filtered data
cat("\nThe following data were generated witht the script 'data_sim_ph_driven.R':\n")

# brings into scope pre-generated data
not_filt_data <- readRDS(
  list.files(
    path = here(),
    pattern = "^data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)


# print the pre-generated data
print(not_filt_data)


# -------- show filtered data --------

# brings into scope pre-generated and then filtered data
filtered_data <- readRDS(
  list.files(
    path = here(),
    pattern = "^filtered_data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

cat("\nThese were the unfiltered data.\n")
cat("\nNow we will apply the filt_data() function with a threshold of 0.3, obtaining the
    following data:\n")

# print the generated and filtered data
print(filtered_data)

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

### pseudocount.R

[*What it does:*]{style="color:green"} defines the function `pseudocount()`, which converts count data to row-wise proportions and replaces zeroes with row-specific pseudocounts computed from a user-defined threshold.

[*Why is it here:*]{style="color:green"} before applying the Centered Log-Ratio (CLR) transformation, it is necessary to substitute the zeroes in the dataset with pseudocounts (log(0) gives −inf as a result, *kinda* difficult to use for statistical analysis).

[*How it works:*]{style="color:green"}

- the script first asks the user for a number between 0 and 1 (if the user does not provide a valid one, the script keeps asking for it), that will be used as threshold;

- it computes the detection limit as 1 / (the sum of the values in the row);

- the pseudocount is computed as the detection limit times the threshold provided by the user;

- it converts every row's counts into proportions;

- it substitutes every zero found in a row with the pseudocount for that row.

[*Note:* due to the interactive nature of the pseudocount.R script (the script asks the user for a number between 0 and 1 which is used as threshold to set the filter) , the following data has been generated with the script `for_notebook.R`, then loaded here.]{style="color:red"}

```{r}
#| echo: false

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")

# tell the user that the example matrix used before will be printed again
cat("\nNow we will print the example_matrix that we used before:\n")

# print the example matrix
print(example_matrix)

# tell the user that the matrix passed thorugh the pseudocount() function will be printed
cat("\nAfter using the pseudocount() function with 0.5 as threshold,
    the following matrix is obtained:\n")

# this matrix was computed outside the notebook and then hardcoded here to avoid
# to launch the pseudcocunt.R script that might have caused the chunk to freeze
pseudocounted_matrix <- matrix(
  c(0.759, 0.241, 0.0172, 0.25, 0.75, 0.0417, 0.312, 0.625, 0.0625),
  nrow = 3, ncol = 3,
  dimnames = list(c("alfa", "beta", "gamma"), c("A", "B", "C"))
)
# print the pseudocounted matrix
print(pseudocounted_matrix)

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

### generate_matrix_factors.R

[*What it does:*]{style="color:green"} defines the function `generate_matrix_factors()`, which builds a valid correlation matrix (symmetric, unit diagonal, positive semi-definite, values from -1 to +1) with an exact, controlled zero pattern, by splitting taxa into `n_groups` latent groups.

[*Why is it here:*]{style="color:green"} in the data_sim_ph_driven.R script is used the function "rmvnorm" ("Random MultiVariateNORMal") which allows to generate random multivariate normal distributions.\
One of the input of the "rmvnorm" function is the correlation matrix.
Utilizing this script allows to have a controlled (and yet partially randomized)
correlation matrix.

[*How it works:*]{style="color:green"} `generate_matrix_factors(n, n_groups, ...)`:

- randomly assigns each of the `n` taxa to one of `n_groups` groups;
- builds a loadings matrix $\Lambda$ where each taxon has a random value only in the column of its own group;
- computes $\Sigma = \Lambda\Lambda'$ (always PSD, whatever $\Lambda$ is);
- rescales it to a proper correlation matrix (unit diagonal);
- returns the matrix, the group assignment, and the exact number of zero pairs (taxa in different groups end up with correlation exactly 0).

```{r}
# brings the generate_matrix_factors script into scope
source(
  list.files(
    path = here(),
    pattern = "^generate_matrix_factors\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# generate a demostrative matrix with n = 10, n_groups = 2, seed = 42
corr_mat_notebook <- generate_matrix_factors(n = 10, n_groups = 3, seed = 42)

# print the actual matrix
print(corr_mat_notebook$mat)

# print a correlation matrix with no OTU correlated
not_corr_mat_notebook <- generate_matrix_factors(n = 10, n_groups = 10, seed = 42)

# print the non-correlation matrix
print(not_corr_mat_notebook$mat)
```

### pielou_ind.R

[*What it does:*]{style="color:green"} defines the function `pielou_ind()`, which computes the dataset diversity as the mean of the pielou indexes of all the samples in the dataset.

[*Why is it here:*]{style="color:green"} it is a good parameter (used also in the original paper) that allows to quantify the species diversity of the sample.\
It is needed for the final analysis

[*How it works:*]{style="color:green"} `pielou_ind(x, per_sample = FALSE)`:

- computes each sample's relative abundances and Shannon entropy, handling zeroes with the standard convention $0 \cdot \ln(0) = 0$, so it works directly on sparse real data, no pseudocounts needed;
- normalizes by $\ln(D)$ (D = number of taxa), so the index stays comparable across datasets of different size;
- returns the mean across samples by default, or the full per-sample vector if `per_sample = TRUE`.

```{r}
#| echo: false
# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")


# -------- sourced stuff needed for this chunk --------

# brings the pielou_ind script into scope
source(
  list.files(
    path = here(),
    pattern = "^pielou_ind\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# read previously generated data
data_for_notebook_count <- readRDS(
  list.files(
    path = here(),
    pattern = "^data_for_notebook_count\\.rds$",
    full.names = TRUE,
    recursive = TRUE
  )
)

# -------- display of script activity --------

# shows the sourced data
print(data_for_notebook_count)

# explain the user what's happening
cat("\nApplying the pielou_ind() function to the dataset
(using per_sample = FALSE),
we obtain the within dataset diversity:\n")

# apply the pielou_ind() function to the sourced data (per_sample = FALSE will
# give the within dataset diversity parameter):
round(pielou_ind(data_for_notebook_count, per_sample = FALSE), 2)

# explain the user what's happening
cat("\nNow, if we use the same function with per_sample = TRUE,
we obtain a vector with all of the pielou indexes of the
data (data has 8 sample, which means 8 rows, so we will
obtain 8 indexes):\n")

# apply the pielou_ind() function to the sourced data using per_sample = TRUE
pi_ind_vect <- pielou_ind(data_for_notebook_count, per_sample = TRUE)

# show the various pielou index comouted from the dummy dataset
print(pi_ind_vect)

# explain the user what's happening
cat("\nIf we compute the mean of the previously generated vector,
we obtain again the within dataset diversity parameter:\n")

# show that applying the mean to the pi_ind_vect gives us the already printed within
# dataset diversity parameter
print(round(mean(pi_ind_vect), 2))

# spacing bar
cat("\n#-----------------------------------------------------------------------#\n")
```

## Orchestrators

Orchestrators are the next level of base functions: instead of doing a single small job, they combine several base functions together into a full step of the analysis (filtering + pseudocount + CLR + correlation, or building a correlation matrix + simulating data from it).

### clr_pearson.R

[*What it does:*]{style="color:green"}

[*Why is it here:*]{style="color:green"}

[*How it works:*]{style="color:green"}

```{r}
# brings the clr_pearson script into scope
source(
  list.files(
    path = here(),
    pattern = "^clr_pearson\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

### NorTa_simulation.R

[*What it does:*]{style="color:green"}

[*Why is it here:*]{style="color:green"}

[*How it works:*]{style="color:green"}

```{r}
# brings the NorTa_simulation script into scope
source(
  list.files(
    path = here(),
    pattern = "^NorTa_simulation\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

### data_sim_ph_driven.R

[*What it does:*]{style="color:green"}

[*Why is it here:*]{style="color:green"}

[*How it works:*]{style="color:green"}

```{r}
# brings the data_sim_ph_driven script into scope
source(
  list.files(
    path = here(),
    pattern = "^data_sim_ph_driven\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

## Standalone analysis

The final level of complexity of all the project.  
The scripts presented in this section perform the main statistical analyses of the project.   
Each analysis is designed to investigate a specific issue identified in the original paper and to assess its potential implications for the results.

### compositional_bias_D_P.R

[*What it does:*]{style="color:green"}

[*Why is it here:*]{style="color:green"}

[*How it works:*]{style="color:green"}

```{r}
# brings the compositional_bias_D_P script into scope
source(
  list.files(
    path = here(),
    pattern = "^compositional_bias_D_P\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```

### ph_pielou_confrontation.R

[*What it does:*]{style="color:green"} uses `data_sim_ph_driven()` across a pH gradient, at five different levels of pH tolerance, and tracks how the Pielou index changes, showing that narrow-tolerance ("specialist") communities produce noisier, lower evenness, while wide-tolerance ("generalist") communities stay stable and high.

[*Why is it here:*]{style="color:green"} This is the final exame of this project. I needed something that could test how data_sim_ph_driven.R could have been employed (in this case to simulate how generalists vs specialists bacteria behave at the same pH).

[*How it works:*]{style="color:green"} loops over a pH gradient for five tolerance scenarios, computes `pielou_ind()` at each point, plots the five curves, and combines them into a single figure.

```{r}
# brings the ph_pielou_confrontation script into scope
source(
  list.files(
    path = here(),
    pattern = "^ph_dataset_diversity_confrontation\\.R$",
    full.names = TRUE,
    recursive = TRUE
  )
)
```
